Assessing the Feasibility of a Cluster Randomised Trial on the Effect of Trauma Life Support Training: A Pilot Study

### 2024-11-02 pilot-publication/manuscript.Rmd
Add new citations

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

@@

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm. The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS and PTC arms respectively, as shown in Table \@ref(tab:outcomes). Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS and PTC arms were very small. The results for all other outcomes are shown in Supplementary Tables S1-12.

@@

Previous studies on the effect of in-hospital trauma life support training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008; @Wang2010]. Studies from Trinidad and Tobago, El Salvador, Rwanda, and Cambodia found no significant effect on patient mortality after implementing in-hospital trauma life support training [@Ariyanayagam1992; @CioèPeña2016; @Petroze2014; @VanHeng2008], whereas one study from China that included 820 patients found a significant reduction in mortality, from 20 to 15%, after implementing ATLS^®^ [@Wang2010].

>>>>

The proprietary Advanced Trauma Life Support^®^ (ATLS^®^) and the low-cost alternative Primary Trauma Care (PTC) are two widely established trauma life support training programmes with over a million physicians trained in over 80 countries [@acsAtls2018; @ptcfPtc2018]. Observational studies indicate that these programmes may improve patient outcomes [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016], but there is no high quality evidence from controlled trials to support this [@Mohammad2013; @Jayaraman2014; @Kadhum2020; @Jin2021; @putra_impact_2023; @mciver_effect_2024].

@@

After training, a total of `r m30d.standard.care` patients in the standard care arm died within 30 days, compared to `r m30d.atls` patients in the ATLS^®^ arm and `r m30d.ptc` patients in the PTC arm. The corresponding figures for in-hospital mortality were `r inhosp.standard.care`%, `r inhosp.atls`%, and `r inhosp.ptc`% for the standard care, ATLS^®^ and PTC arms respectively, as shown in Table \@ref(tab:outcomes). Overall, both in-hospital and 30-day mortality were substantially lower in the ATLS^®^ and PTC arms compared to the standard care arm, but the absolute numbers of deaths in the ATLS^®^ and PTC arms were very small. The results for all other outcomes are shown in Supplementary Tables S1-12.

@@

Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016]. In contrast, some studies have found that these programmes are associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016].
